---
title: 'ARIMA forecasting - Spain metro users'
author: "Javier Eloy Martinez Ramos"
date: "3/3/2022"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r include=FALSE}
library(readxl)
library(lubridate)
library(ggplot2)
library(dplyr)
library(zoo)
library(forecast)
library(kableExtra)
library(tseries)
```

## Introduction

This dataset was downloaded from the spanish National Institute of Statistics
(INE) and it contains the quantities of metro users in Spain, from january 2011
to december 2019.

```{r include=FALSE}
tabla <- read_excel("metroEspana.xlsx")

datos <- tabla[8:115,]
colnames(datos)[2] <- "Viajeros"

viajeros <- ts(datos[, 2], start = c(2011, 1), frequency = 12)
```

## Graphic representation and seasonal decomposition

```{r echo=TRUE}
autoplot(viajeros) + 
  ggtitle("Metro travelers in Spain") + 
  xlab("Date") + ylab("Number of travelers")
```

Here we can see that the data doesn't move too much, this can mean that our 
serie is stationary.

```{r echo=TRUE}
viajeros_Comp <- decompose(viajeros, type = c("multiplicative"))

autoplot(viajeros_Comp, ts.colour = "blue")
```

Here we can see that the trend does move, and we clearly have seasonality.

Then we represent the serie with trend and the seasonally adjusted serie:

```{r echo=TRUE}
autoplot(viajeros, series = "Data") +
  autolayer(trendcycle(viajeros_Comp), series = "Trend") +
  autolayer(seasadj(viajeros_Comp), series = "Seasonally adjusted") +
  xlab("Date") + ylab("Number of travelers") +
  ggtitle("Metro travelers in Spain") +
  scale_colour_manual(
    values = c("#736F4E", "#4C061D", "#D17A22"),
    breaks = c("Data", "Seasonally adjusted", "Trend")
    )
```

Then we see the seasonal representation:

```{r echo=TRUE}
seasonplot <- ggseasonplot(viajeros, year.labels = TRUE, year.labels.left = TRUE) +
  ylab("Number of travelers") + xlab("Date") +
  ggtitle("Seasonal plot: Metro travelers in Spain")

seasonplot$labels$group <- "Year"
seasonplot$labels$colour <- "Year"

seasonplot
```

There are similarities between years in terms of shape, but not too evident.

## Dataset partition

In order to check the accuracy of the forecast methods we are partitioning the
dataset to compare forecast and actual events.

```{r echo=TRUE}
# We leave out the data corresponding to the last 12 months.

reservados <- 12

viajeros_mod <- viajeros[1:(nrow(viajeros) - reservados),]
viajeros_test <- viajeros[(nrow(viajeros) - reservados):(nrow(viajeros)),]

viajeros_mod <- ts(viajeros_mod, start = c(2011, 1), frequency = 12)
```

## Search for the right time series behaviour model

We chose the Holt-Winters multiplicative model since it's better suited for 
seasonal series.

```{r echo=TRUE}
fit1 <- hw(viajeros_mod, h = reservados, seasonal = "multiplicative", level = c(80, 95))

autoplot(fit1) +
  ggtitle("Holt-Winters method forecast") +
  ylab("Number of travelers") + xlab("Time")
  
autoplot(viajeros) +
  ggtitle("Actual data") +
  ylab("Number of travelers") + xlab("Time")
```

Then we are going to try to obtain a better forecast using ARIMA models.

## Correlogram representation

First, we are going to adjust the right model while also checking that its 
residuals are not correlated.

```{r echo=TRUE}
# We make the Dickey-Fuller test to see if we in fact have a seasonal serie:
adf.test(viajeros_mod, alternative = "stationary")
```

The Dickey Fuller test tells us that the serie we have is seasonal (the P-value 
is smaller than 0.05), so we don't need to adjust the serie. (0 differences)

```{r echo=TRUE}
# ARIMA: (Autoregresivo, diferencias, medias móviles)

# Moving averages:
ggAcf(viajeros) + ggtitle("Moving averages")
# Autoregressives:
ggPacf(viajeros_mod) + ggtitle("Autoregressives")
```

We can see that we have 1 autoregressive and 1 moving average. This means our 
ARIMA model is (1, 0, 1)(1, 0, 0)[12].

```{r echo=TRUE}
modelo <- arima(viajeros_mod, order = c(1, 0, 1), seasonal = c(1, 0, 0))
modelo
checkresiduals(modelo)
```

The P-value of the Ljung-Box test is bigger than 0.05, and this means that the
model is well-adjusted.

Also, we can see in the residuals graphic that we have a pattern that is similar 
to white noise, this means we have no correlation between residuals.

## Diagnosis:

```{r echo=TRUE}
error = residuals(modelo)
# We check that the average of the error is near zero:
autoplot(error)
```

Graphically, the average has the appearance of being zero.

## Forecasting with ARIMA:

```{r echo=TRUE}
pronostico <- forecast(modelo, h = reservados)
pronostico
autoplot(pronostico)
autoplot(viajeros)
```

## Conclusion:

Through the ARIMA model, we have forecasted a similar pattern to the actual 
data, and even if the amplitude of the variations is not exact, the waveshape 
and frequency are very similar, and the actual data is between the marked error
margins.